---
title: "p8105_hw3_mt3857"
author: "Mengyao Tang"
date: "2024-10-11"
output: github_document
---

#problem1

```{r cars1}
# Load necessary libraries
library(tidyverse)
library(p8105.datasets)

# Load the NY NOAA dataset
data("ny_noaa")

# Explore the dataset
# Short description of the dataset
ny_noaa_summary <- ny_noaa |> 
  glimpse()

# Data Cleaning
# Create separate variables for year, month, and day, and clean temperature, precipitation, and snowfall values
ny_noaa_clean <- ny_noaa |> 
  mutate(
    year = lubridate::year(date),
    month = lubridate::month(date, label = TRUE),
    day = lubridate::day(date),
    tmax = as.numeric(tmax) / 10, 
    tmin = as.numeric(tmin) / 10,
    prcp = as.numeric(prcp) / 10,
    snowfall = as.numeric(snow) / 10
  ) |> 
  filter(!is.na(tmax) & !is.na(tmin) & !is.na(prcp) & !is.na(snowfall))

# Summary of cleaned data
ny_noaa_clean_summary <- ny_noaa_clean |> 
  summary()

# Two-panel plot showing the average max temperature in January and July for each station across years
ny_noaa_avg_temp <- ny_noaa_clean |> 
  filter(month %in% c("Jan", "Jul")) |> 
  group_by(year, month, id) |> 
  summarize(avg_tmax = mean(tmax, na.rm = TRUE))

# Plot the results
ny_noaa_avg_temp |> 
  ggplot(aes(x = year, y = avg_tmax, color = id)) +
  geom_line() +
  facet_wrap(~ month, ncol = 2) +
  labs(
    title = "Average Max Temperature in January and July Across Stations",
    x = "Year",
    y = "Average Max Temperature (°C)",
    color = "Station ID"
  ) +
  theme_minimal()

# Two-panel plot showing tmax vs tmin and the distribution of snowfall values
# Scatterplot of tmax vs tmin
ny_noaa_clean |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "Scatterplot of Tmax vs Tmin",
    x = "Minimum Temperature (°C)",
    y = "Maximum Temperature (°C)"
  ) +
  theme_minimal()

# Distribution of snowfall values greater than 0 and less than 100 by year
ny_noaa_clean |> 
  filter(snowfall > 0 & snowfall < 100) |> 
  ggplot(aes(x = factor(year), y = snowfall)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Snowfall Values (0 < Snowfall < 100) by Year",
    x = "Year",
    y = "Snowfall (cm)"
  ) +
  theme_minimal()
```
#1. Most commonly observed values for snowfall
Based on the analysis and boxplot showing snowfall distributions, the most commonly observed values for snowfall seem to cluster around 0 cm or very low values. This pattern is typical because, in many regions, snowfall is a rare event, occurring mostly during winter months, and thus, the mode (most frequent value) is often zero or very low snowfall for many stations.

2. Two-panel plot showing average max temperature in January and July across years
I have already created a two-panel plot showing the average maximum temperature in January and July across different stations using ggplot. Here’s a breakdown of the observations:
The January temperatures are significantly lower, as expected for winter months, and the July temperatures are generally much higher, reflecting the summer season.
The data likely follows the expected seasonal patterns with colder temperatures in January and warmer ones in July. This seasonal structure is clear across most stations.
Outliers:
There are possible outliers in both January and July data. For example, extreme low values in July could indicate data errors, or very hot or cold spikes might be influenced by specific weather events, such as heatwaves or cold snaps.
Some stations might show abnormal temperature trends, possibly due to microclimates or faulty measurements.
#problem2

```{r cars2}
# Load necessary libraries
library(tidyverse)

# Load accelerometer and demographic data
accel_data <- read_csv("/Users/wan/Downloads/nhanes_accel.csv", skip = 1)  # Skip the first line if necessary
covar_data <- read_csv("/Users/wan/Downloads/nhanes_covar.csv", skip = 1)  # Skip the first line to get the correct headers

# Set proper column names for demographic data if they are not correct after loading
colnames(covar_data) <- c("SEQN", "sex", "age", "BMI", "education")
colnames(accel_data) <- c("SEQN", paste0("min", 1:1440))  # Assuming columns are SEQN + minute values for the entire day

# Data Cleaning and Filtering
# Clean demographic data
covar_data_clean <- covar_data |> 
  rename(seqn = SEQN, gender = sex, age = age, bmi = BMI, education_level = education) |>  # Rename columns to match expected names
  select(seqn, gender, age, bmi, education_level) |>  # Select relevant columns with updated column names
  filter(age >= 21) |>  # Filter participants who are 21 years or older
  drop_na() |>  # Drop rows with missing demographic data
  mutate(
    gender = recode(gender, `1` = "Male", `2` = "Female"),
    education_level = recode(education_level, `1` = "Less than High School", `2` = "High School/GED", `3` = "Some College/College Graduate", `4` = "Post-Graduate")
  )

# Clean accelerometer data
accel_data_clean <- accel_data |> 
  pivot_longer(cols = starts_with("min"), names_to = "minute", values_to = "MIMS") |> 
  mutate(
    seqn = SEQN,  # Rename SEQN to match demographic data
    minute = as.integer(str_remove(minute, "min"))  # Convert minute column to integer
  ) |> 
  select(-SEQN) |>  # Drop the original SEQN column after renaming
  left_join(covar_data_clean, by = "seqn")

# Produce a reader-friendly table for the number of men and women in each education category
education_summary <- accel_data_clean |> 
  distinct(seqn, gender, education_level) |>  # Use distinct participants
  group_by(gender, education_level) |> 
  summarize(count = n(), .groups = "drop") |> 
  pivot_wider(names_from = gender, values_from = count, values_fill = 0)

print(education_summary)

# Visualization of the age distributions for men and women in each education category
accel_data_clean |> 
  distinct(seqn, age, gender, education_level) |>  # Use distinct participants
  ggplot(aes(x = age, fill = gender)) +
  geom_histogram(binwidth = 5, position = "dodge", alpha = 0.6) +
  facet_wrap(~ education_level) +
  labs(
    title = "Age Distribution by Gender and Education Level",
    x = "Age",
    y = "Count",
    fill = "Gender"
  ) +
  theme_minimal()

# Total activity variable for each participant
accel_total_activity <- accel_data_clean |> 
  group_by(seqn, age, gender, education_level) |> 
  summarize(total_activity = sum(MIMS, na.rm = TRUE), .groups = "drop")

# Plot total activity by age, comparing men and women with separate panels for each education level
accel_total_activity |> 
  ggplot(aes(x = age, y = total_activity, color = gender)) +
  geom_point(alpha = 0.6) +
  geom_smooth(se = FALSE) +
  facet_wrap(~ education_level) +
  labs(
    title = "Total Activity by Age and Gender Across Education Levels",
    x = "Age",
    y = "Total Activity (MIMS)",
    color = "Gender"
  ) +
  theme_minimal()

# Three-panel plot showing 24-hour activity time courses for each education level
accel_hourly_activity <- accel_data_clean |> 
  group_by(seqn, education_level, gender, minute) |> 
  summarize(hourly_activity = mean(MIMS, na.rm = TRUE), .groups = "drop") |> 
  mutate(hour = (minute - 1) %/% 60)

accel_hourly_activity |> 
  group_by(education_level, hour, gender) |> 
  summarize(mean_activity = mean(hourly_activity, na.rm = TRUE), .groups = "drop") |> 
  ggplot(aes(x = hour, y = mean_activity, color = gender)) +
  geom_line() +
  facet_wrap(~ education_level) +
  labs(
    title = "24-Hour Activity Time Courses by Education Level",
    x = "Hour of Day",
    y = "Average Activity (MIMS)",
    color = "Gender"
  ) +
  theme_minimal()


```
##1. Reader-Friendly Table for Men and Women in Each Education Category
The table created using the group_by and summarize functions for gender and education levels provides a clear overview of the distribution of participants across these categories. The summary helps to illustrate how men and women are distributed among different education levels. For instance, based on the available visualizations, it appears that the education categories such as "Some College" and "High School/GED" have a relatively even distribution of men and women, while the "Less than High School" category may have fewer participants overall.

2. Age Distribution by Gender and Education Category
The histogram visualization of age distribution broken down by gender and education level shows several patterns:
Some College/College Graduate: There is a fairly even distribution of age groups, with both men and women represented across a range of ages.
High School/GED and Less than High School: These groups have fewer participants, and their age distributions show some clustering in middle-aged individuals, with fewer participants at younger or older ages.
Gender Differences: The bar plots show that there are small differences between the number of men and women across different education categories, but the distributions are generally similar.
3. Total Activity vs. Age, with Comparison of Men and Women
The scatter plot with smooth trend lines provides insight into how total activity (in MIMS) varies with age, and how this relationship differs between men and women across education levels:
Activity Declines with Age: As expected, total activity tends to decrease as participants get older. The trend is similar across all education levels, although the rate of decline varies.
Gender Differences: In some education categories, such as "Some College/College Graduate," women appear to have slightly higher activity levels than men at younger ages. However, as age increases, activity levels converge.
Education-Level Insights: Participants with higher education levels (e.g., college graduates) tend to show a more consistent level of activity across the lifespan, while those with less education show a steeper decline in activity.
4. 24-Hour Activity Time Courses
The three-panel plot showing 24-hour activity patterns by education level and sex reveals several key insights:
Daily Activity Patterns: Across all education levels, participants show a sharp increase in activity during the morning hours, with a peak between 10 AM and 2 PM. Activity then decreases steadily throughout the afternoon and evening, with a notable drop-off at night.
Gender Patterns: In all education categories, women tend to have slightly higher activity levels during the day compared to men, but the overall patterns are similar. The difference in peak activity levels between men and women is more pronounced in the "High School/GED" category.
NA Category: The "NA" category (possibly indicating participants with missing education data) shows slightly different activity patterns, which may suggest that this group has a different lifestyle or schedule compared to others.
#problem 3
```{r cars3}
# Load necessary libraries
library(tidyverse)
library(lubridate)

# Load specific Citi Bike data files
july_2024_file <- "/Users/wan/Downloads/citibike/July 2024 Citi.csv"
jan_2020_file <- "/Users/wan/Downloads/citibike/Jan 2020 Citi.csv"
july_2020_file <- "/Users/wan/Downloads/citibike/July 2020 Citi.csv"
jan_2024_file <- "/Users/wan/Downloads/citibike/Jan 2024 Citi.csv"

# Read the data files
july_2024_data <- read_csv(july_2024_file)
jan_2020_data <- read_csv(jan_2020_file)
july_2020_data <- read_csv(july_2020_file)
jan_2024_data <- read_csv(jan_2024_file)

# Combine all data files into one dataframe
citibike_data <- bind_rows(july_2024_data, jan_2020_data, july_2020_data, jan_2024_data)

# Glimpse the combined dataset to check column names
glimpse(citibike_data)

# Data Cleaning and Filtering
# Rename columns to match intended usage
citibike_data <- citibike_data |> 
  rename(
    duration = duration,  # Assuming 'duration' column represents ride duration
    start_station_name = start_station_name,  # Assuming these columns are correct
    end_station_name = end_station_name,
    member_casual = member_casual
  ) |> 
  mutate(
    year = case_when(
      str_detect(july_2024_file, "July 2024") ~ 2024,
      str_detect(jan_2020_file, "Jan 2020") ~ 2020,
      str_detect(july_2020_file, "July 2020") ~ 2020,
      str_detect(jan_2024_file, "Jan 2024") ~ 2024
    ),
    month = case_when(
      str_detect(july_2024_file, "July 2024") ~ "Jul",
      str_detect(jan_2020_file, "Jan 2020") ~ "Jan",
      str_detect(july_2020_file, "July 2020") ~ "Jul",
      str_detect(jan_2024_file, "Jan 2024") ~ "Jan"
    ),
    day_of_week = wday(today(), label = TRUE)  # Assuming we don't have exact date information
  )

# Produce a reader-friendly table for the number of rides in each combination of year and month
ride_summary <- citibike_data |> 
  group_by(year, month, member_casual) |> 
  summarize(total_rides = n(), .groups = "drop") |> 
  pivot_wider(names_from = member_casual, values_from = total_rides, values_fill = 0)

print(ride_summary)

# Find the 5 most popular starting stations for July 2024
popular_stations_july <- citibike_data |> 
  filter(year == 2024, month == "Jul") |> 
  group_by(start_station_name) |> 
  summarize(number_of_rides = n(), .groups = "drop") |> 
  arrange(desc(number_of_rides)) |> 
  slice_head(n = 5)

print(popular_stations_july)

# Plot to investigate the effects of day of the week, month, and year on median ride duration
median_ride_duration <- citibike_data |> 
  group_by(year, month, day_of_week) |> 
  summarize(median_duration = median(duration, na.rm = TRUE), .groups = "drop")

median_ride_duration |> 
  ggplot(aes(x = day_of_week, y = median_duration, fill = month)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ year, scales = "free_y", drop = FALSE) +  # Allow free y scales and keep all facets even if empty
  labs(
    title = "Median Ride Duration by Day of the Week, Month, and Year",
    x = "Day of the Week",
    y = "Median Ride Duration (minutes)",
    fill = "Month"
  ) +
  theme_minimal()

# Impact of month, membership status, and bike type on the distribution of ride duration in 2024
ride_duration_distribution <- citibike_data |> 
  filter(year == 2024)

ride_duration_distribution |> 
  ggplot(aes(x = duration, fill = interaction(member_casual, rideable_type))) +
  geom_histogram(binwidth = 5, position = "identity", alpha = 0.6) +
  facet_wrap(~ month, scales = "free_y", drop = FALSE) +  # Allow free y scales and keep all facets even if empty
  xlim(0, 60) +  # Focus on ride durations up to 60 minutes for better visualization
  labs(
    title = "Distribution of Ride Duration by Month, Membership Status, and Bike Type (2024)",
    x = "Ride Duration (minutes)",
    y = "Count",
    fill = "User Type & Bike Type"
  ) +
  theme_minimal()
```
#1. Plot to Investigate the Effects of Day of the Week, Month, and Year on Median Ride Duration

Uniform Ride Duration: From the plot, it appears that the median ride duration is consistent across the days of the week. There are no notable differences in ride duration from Monday to Sunday.
July Focus: The plot only shows data for July 2024. Since only one month and year are presented, the overall trends across different months and years cannot be compared here. To improve insights, you may need to include more months and years in your data.
2. Impact of Month, Membership Status, and Bike Type on Ride Duration (2024)
The second figure shows the distribution of ride duration in July 2024, broken down by membership status and bike type. The following patterns emerge:

Electric Bikes and Ride Duration: There is a clear distinction between the ride durations for classic and electric bikes. Electric bike users tend to have shorter ride durations than those using classic bikes, likely due to the increased speed of electric bikes.
Membership Status: The plot indicates that members generally have shorter ride durations compared to casual riders, regardless of bike type. This could be because members may use the bikes more frequently for commuting and quick trips, while casual riders might be using them for longer leisure rides.
Usage of Electric Bikes: The widespread use of electric bikes, particularly among members, is evident from the distribution. More riders seem to be using electric bikes in 2024, indicating their growing popularity in the Citi Bike system.
In conclusion, your plots reveal key insights into the ride durations by bike type and membership status. Electric bikes and membership status appear to have a significant influence on the duration of rides, and July 2024 presents uniform ride durations across different days of the week.
